---
title: "test"
output: html_document
---

## Packages

```{r}
library(wooldridge) #download database
library(tidyverse)c
library(stargazer) #to draw outpout tables
library(lmtest) #to test linear progressions models
library(AER) #contains some usefull econometrics functions in R
```

##we want to explain wage with others variables
##so we load the data

```{r}
data("wage2") #to load the data

head(wage2, n=20) #to see the first 20 entries
```
```{r}
summary(wage2)
```

```{r}
summary(wage2$educ) 
```
## summary() : Min, Max, ...

#How many people live in a Urban area ?
## table() : gives the frequency


```{r}
table(wage2$urban)
```

```{r}
# R syntax :
table(wage2$married, wage2$urban, dnn = c("married", "urban"))

#tidyverse syntax 
wage2 %>%
  count(married, urban)
```

#what is the average wage for married people ?
#agregate()
```{r}
#R syntax :
aggregate(wage~married, data = wage2, FUN=mean)

##tidyverse syntax :
wage2 %>%
  group_by(married) %>%
  summarise(mean(wage))

```

# What is the correlation between wage and education ? Is this correlation significant ?
#cor() = correlation
```{r}

cor(wage2$wage, wage2$educ)
#is the correlation significant ?
cor.test(wage2$wage, wage2$educ)

#confidence interval : voir la significance dans le cours, proche de 0 plus y a de la correlation
#signficance: proof your hypothesis (like this variable explains this variable)
```
# estimate a regression model ...
#lm()
#name of the model<-lm(dependent~indep1+indep2..., data = wage2)
```{r}
m1<- lm(wage ~ IQ, data = wage2)

#R syntax
summary(m1) # to check what my model 1 gives

#option 2 : show results with stargaze : pas eu le temps de noter voir correc
stargazer(m1, type = "text")

#prédire pour an increase de l'IQ de 15 points :
#estimate = value which explain others (B coefficient)
#pr = signifance (p-value), here its close to 0, so IQ is relevant
#wage = B0(=intercept) + B1(IQ = 8.3) + e(epsilon)
#if IQ increases by 1 unit => my wage increases by 8.3
#Signif codes= from 0 to 0.001 its ***, from ...


15*8.3 #124.5 dollars increase in wage for an increase of 15 pts of IQ

#Multiple R-squared = Is my model is well explaining ? y a d'autres variables qui peuvent expliquer wage. Donc a quel point IQ the closer is to 1 so more my value explique bien? Plus c proche de 1 plus ca explique bien.
#significance = c'est different de multiple R-squared parce que significance c'est a quel point c'est relevant la correlation entre 2 variables
```
```{r}
#log(wage) : transforme en pourcentage
wage2$lwage <- log(wage2$wage) #creating a new variable lwage et ensuite affection des valeurs en pourcentage de wage, dans une nouvelle colonne lwage qu'on crée


m2<- lm(lwage ~ IQ, data = wage2) #on crée le model m2
stargazer(m2, type = "text") # show results
```
```{r}

#voir le document quelle a montré avec une grille selon les types de regressions, comment on interprète
m3<- lm(lwage ~ IQ + educ+exper+black+tenure+married+south, data = wage2)
stargazer(m3, type = "text") # show results
```
## On passe au IV

```{r}
#Run the model that estimates the return to education for men
# explain wage by education

m4<- lm(lwage ~ educ, data = wage2)
stargazer(m4, type = "text") # show results

#is there a possible source of endogenity : education can be influenced by ability, ou l'IQ
#an instrument : variable that explains endogenious variables. Here we will try to explain education
#Is there a variable that explains education but not explainging wage : for exemple education of parents : influence wage threw education. Number of sibligns can also be an IV. A lot of siblings, less chances to long education

```
```{r}
#We have to test 2 criterias : sibs correlated to educ ?
# Positive or neg corre between IV and endogenous variables
# IV doesnt explain dependent var
  

#first stage : sibs explain education
# second stage : education will explain wage
  
#sib is negatibly correlated to education

#lm() = was for OLS
#ivreg() = is for IV
#ivreg(dependent vari ~ independent var) | instrument, data = name of data)
# avant le pipe cest le modele que je veux estimer : genre educ + south + black
# mais educ est endogene, mais peut etre explique par sibs
# so wage va etre expliqué par ca + après le pipe je vais mettre le IV model : sib + south + black
# si la variable endogene est a la deuxieme place, faut mettre le IV à la deuxieme place derriere le pipe.
#ivreg(dependent vari ~ independentvar + identvar2(endogene) + indentvar3) | independentvar + instrument + indentvar3, data = name of data)
ivmodel <- ivreg(lwage ~ educ | sibs, data = wage2)
summary(ivmodel)
stargazer(ivmodel, type = "text")
```

```{r}

#refaire la wage, avec en variable endogene education, avec birth order comme IV
ivmodel2 <- ivreg(lwage ~ educ | brthord, data = wage2)
summary(ivmodel)
stargazer(ivmodel2, type = "text")
```
```{r}
#regresser lwage avec educ south, married, experience correcting for endogene educ by using brthord

#ivmodel3 <-ivreg(lwage ~ married + educ + south + experience) | married + brthord + south + exper, data = wage2) jsp pk ca marche pas

ivmodel2 <- ivreg(lwage~educ+south+married+exper|brthord+south+married+exper, data=wage2)
stargazer(ivmodel2, type = 'text')

summary(ivmodel)
stargazer(ivmodel3, type = "text")
```

